Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-08 analysis/paper/paper.Rmd
incorporate AGH suggestions

---
title: "Investigating Biases Associated with Dietary Starch Incorporation into an Oral Biofilm Model"
author:
  - Bjørn Peare Bartholdy
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
#cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
## end refs.bib ##
```

@@

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus, mineralised dental plaque, has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculusCooking2014]. 
This dietary information can be preserved within calculus over many millenia,
providing a unique window into the food-related behaviours of past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has been, 
to a larger extent, acknowledged. The applications of dental calculus span a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsightsConsumption2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored the how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and you have a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus in a controlled laboratory 
setting, allowing us full control over dietary input. Our main questions concern
the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of 
diet from extracted microremains?  
We find that, despite the absence of $\alpha$-amylase in
the model, a limited proportion of the starch input is actually 
retained in the calculus, as well as a shift in the size ratios of individual 
starch granules
that are incorporated into the calculus. We also show that the number of starch
granules that are incorporated increases as the size of the calculus deposit 
increases.

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins.

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of the protocol). More detailed protocols
are available on protocols.io<!--insert DOI-->.

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 5, 6, 9, and 12. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
For a more detailed protocol, see <!-- protocols.io DOI -->

@@

All scripts and data used in the analysis are available on OSF 
(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch).
All protocols are available on protocols.io.

@@

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for later
 analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (see Supplementary Material).

@@

```{r sample-prop-tbl}
cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
knitr::kable(perc, col.names = cnames, 
             caption = "The mean percentage of starches from the solutions that 
              were incorported into the samples.")
```

The proportion of total starches from the solutions that were incorporated 
into the samples ranged from 
`r paste(min(perc$total), max(perc$total), sep = " to ")`,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \@ref(tab:sample-prop-tbl)). 
There is an inverse relationship between the absolute starch count in the solutions 
and the proportional incorporation of starches in the biofilm samples, i.e. potato 
had the lowest absolute count in solutions, but the highest proportional 
incorporation, and vice versa for the mixed treatment.

<!-- separated treatments -->

@@

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
pl_cor
```

@@

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc$total), max(perc$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(count_sol$total[4], big.mark = ",")`) 
and potato
(mean = `r format(count_sol$total[3], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc$total), max(perc$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010]. <!-- although this study extrapolated counts to full sample content, which is
rarely done in archaeological studies -->This could, in part, be due to the 
activity of oral amylase, which was absent in our model.

<!--
Given the low proportion of starch granule retention in our dental calculus model,
and the limited difference between treatments, we can conjecture that starch 
availability is not a main factor in starch incorporation. While there was a
proportional incorporation, i.e. more granules in the solution led to more granules
in the calculus, the difference between solutions [a number/percentage] was of a
higher magnitude than the difference between samples <!-- a number/percentage-->

<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardDentalCalculus2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.

<!-- secondary findings -->
We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10 -- 20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
This was especially true for the potato starches, which can reach up to 100 $\mu$m
in diameter, whereas wheat generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004].
The size bias is consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
which found that plants producing starches 10--20 $\mu$m in
size were over-represented; although, the representation of granules larger than
20 $\mu$m is unclear.  
Given the differences in proportional incorporation between wheat and potato,
especially for small and medium granules, it is also possible that the morphology
of granules plays a role. However, the proportional count differences between large 
wheat and potato granules were less pronounced.

<!-- Since those results were obtained -->
<!-- on a modern population of chimps that were only buried for a short period of time,  -->
<!-- degradation was likely limited [@powerChimpCalculus2015]. -->

The taphonomy of starch granules has been addressed previously, and depends on 
burial environment (pH, temperature, moisture content, microorganisms), as well 
as processing of the granules prior to deposition 
[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
The potential for detrimental diagenetic effects on starch granules increases as 
the size and amylose content decreases
[@francoStarchDegradation1992; @haslamDecompositionStarch2004].

<!-- add potential for morphological differences affecting incorporation 
 Medium potato starches had the highest rate of incorporation 
 All large starches had a consistently low rate of incorporation 
 Starch granule adhesion to bacteria and enzymes? Do they have a preference?
 Binding to select species in a biofilm could explain clustering
 Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017-->

@@

Another potentially important factor 
from our results is the size of the calculus deposit, which seems to influence the 
quantity of starches extracted from the calculus, as we found a strong positive 
correlation between biofilm size and retained starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014]. 
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2));
while the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
The lack of a correlation
in archaeological contexts could be due to diagenetic effects. Alternately, the absence of 
amylase activity in our study could be impacting our interpretations.

<!-- How are starch granules trapped in dental calculus? Does anyone actually know? - adhesion to bacteria
- stuck in cracks and crevices
- covered by new layer of plaque

>>>>

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author:
  - Bjørn Peare Bartholdy
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
    extra_dependencies: "subfig"
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
  bookdown::html_document2:
    df_print: kable
    keep_md: true
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.064% to 0.161%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
#cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
## end refs.bib ##

## Word count ##
wd_count <- wordcountaddin::word_count()
```

Word count: `r wd_count`

@@

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculusCooking2014]. 
This dietary information can be preserved within the mineralised dental plaque 
over many millennia, providing a unique window into the food-related behaviours of 
past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncientCalcified2013; @warinnerPathogensHost2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsightsConsumption2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
and nicotine use [@eerkensDentalCalculusSource2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored the how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus in a controlled laboratory 
setting, allowing us full control over dietary input. Our main questions concern
the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of 
diet from extracted microremains?  
We find that, despite the absence of $\alpha$-amylase in
the model, a limited proportion of the starch input is actually 
retained in the calculus. We also observed a shift in the size ratios of individual 
starch granules that are incorporated into the calculus, and that the number of 
incorporated starch granules increases as the size of the calculus deposit 
increases.

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins.

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of the protocol).

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.

@@

All scripts and data used in the analysis are available on OSF 
(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch).
More detailed protocols are available on protocols.io.

@@

```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\\mu$m.", out.width="50%"}
knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
```

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for later
 analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (see Supplementary Material).

@@

```{r sample-prop-tbl}
cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
knitr::kable(perc, col.names = cnames, 
             caption = "The mean percentage of starches from the solutions that 
              were incorporated into the samples.")
```

The proportion of total starches from the solutions that were incorporated 
into the samples ranged from 
`r paste(min(perc$total), max(perc$total), sep = " to ")`,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \@ref(tab:sample-prop-tbl)). 
There is an inverse relationship between the absolute starch count in the solutions 
and the proportional incorporation of starches in the biofilm samples, i.e., potato 
had the lowest absolute count in solutions, but the highest proportional 
incorporation, and vice versa for the mixed treatment.

@@

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc$total), max(perc$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(count_sol$total[4], big.mark = ",")`) 
and potato
(mean = `r format(count_sol$total[3], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc$total), max(perc$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], <!-- although this study extrapolated counts to full sample content, which is
rarely done in archaeological studies -->which could in part be due to the 
lack of oral amylase activity in our model.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2)).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
Further research is needed to determine why this differs from previous archaeological
findings.
<!--
Given the low proportion of starch granule retention in our dental calculus model,
and the limited difference between treatments, we can conjecture that starch 
availability is not a main factor in starch incorporation. While there was a
proportional incorporation, i.e. more granules in the solution led to more granules
in the calculus, the difference between solutions [a number/percentage] was of a
higher magnitude than the difference between samples <!-- a number/percentage-->

<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardDentalCalculus2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
who found that plants producing starches 10--20 $\mu$m in
size were over-represented; however, the representation of granules larger than
20 $\mu$m in their study is unclear.

<!-- add potential for morphological differences affecting incorporation 
 Medium potato starches had the highest rate of incorporation 
 All large starches had a consistently low rate of incorporation 
 Starch granule adhesion to bacteria and enzymes? Do they have a preference?
 Binding to select species in a biofilm could explain clustering
 Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017 (although not
 clear if the bacteria in our model are even 'consuming' starch-->

@@

<!-- How are starch granules trapped in dental calculus? Does anyone actually know? 
-->
The mechanism by which starch granules are incorporated into plaque and calculus 
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering 
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age 
[@henryNeanderthalCalculus2014; @buckleyDentalCalculusCooking2014; @wuDietEarliest2021]. 
Studies have also shown that not all starch granules come from a dietary source. 
Other pathways include cross-contamination from plant interactions in soil, such 
as palm phytoliths adhering to the skin of sweet potatoes 
[@trompDietaryNondietary2015], 
or accidental ingestion not related to food consumption 
[@radiniFoodMultiplePathways2017; @radiniMedievalWomenEarly2019].  
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are 
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to 
tooth surfaces (including plaque and dental calulus) and mucous membranes 
[@kashketFoodRetention1991; @doddsCarbohydrateRetention1988].
Bacteria also have the ability to adhere to starch granules 
[@toppingResistantStarch2003],
